---
title: "Spatial Data Fusion for Extreme Sea Levels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Data Fusion for Extreme Sea Levels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `evfuse` package implements a two-stage frequentist framework for fusing
sparse observations and dense simulations for spatial extreme value analysis
(White et al., in preparation). While developed for U.S. coastal sea levels (NOAA
tidal gauges + ADCIRC simulations), the framework applies to any setting with
annual maxima from multiple spatial data sources.

1. **Stage 1**: Fit GEV distributions independently at each site.
2. **Stage 2**: Fit a joint multivariate Gaussian process that models
   cross-source dependence via a coregionalization (LMC) structure.
3. **Kriging**: Predict at new locations, leveraging information from all
   sources even when each source only observes a subset of the parameters.
4. **Return levels**: Estimate r-year return levels with uncertainty via the
   delta method.

## Data

The package includes `gulf_data`, a dataset of annual maximum sea levels at
129 sites along the U.S. Gulf and Atlantic coasts: 29 NOAA tidal gauge
stations and 100 ADCIRC numerical simulation points.

```{r load-data}
library(evfuse)
data(gulf_data)

gulf_data$n_sites
# [1] 129

head(gulf_data$sites)
```

## Stage 1: GEV Fitting

Fit GEV(mu, sigma, xi) at each site via maximum likelihood. Parameters are
stored as (mu, log sigma, xi) for unconstrained optimization.

```{r stage1}
stage1 <- fit_gev_all(gulf_data)
sum(stage1$converged)
# [1] 129
```

## Bootstrap Measurement Uncertainty

Estimate the covariance of the Stage 1 MLEs via nonparametric block
bootstrap (resampling years to preserve spatial dependence within each data
source). Apply Wendland tapering to enforce sparsity.

```{r bootstrap}
D <- compute_distances(gulf_data$sites)
bs <- bootstrap_W(gulf_data, B = 500, seed = 42)
W_tap <- taper_W(bs$W_bs, D, lambda = 300)
```

## Stage 2: Joint Spatial Model

The joint model has p=6 dimensions (3 per source) with parameters:

- **beta** (6x1): Mean vector for (mu_NOAA, log_sigma_NOAA, xi_NOAA,
  mu_ADCIRC, log_sigma_ADCIRC, xi_ADCIRC).
- **A** (6x6 lower triangular): Cross-covariance factor such that the
  inter-parameter covariance at distance 0 is A A^T.
- **rho** (6x1): Range parameters for the 6 latent exponential
  correlation functions.

The observed parameter vector theta_hat ~ N(beta x 1_L, Sigma + W_tap),
where each site contributes only its observed dimensions (NOAA: 1-3,
ADCIRC: 4-6).

```{r stage2}
model <- fit_spatial_model(stage1, gulf_data, W_tap, D,
                           control = list(maxit = 2000, trace = 0))
model$optim_result$convergence
# [1] 0
```

## Kriging and Return Levels

Predict at new locations. The kriging predictor fuses information from both
NOAA and ADCIRC observations to produce NOAA-scale GEV parameters with
reduced uncertainty.

```{r predict}
grid <- read.csv("data-raw/prediction_grid.csv")
preds <- predict_krig(model, grid)
rl <- compute_return_levels(preds, r = 100)

# 100-year return levels at first 5 grid points
head(rl$return_level, 5)
```

## Model Validation: Leave-One-Out CV

The closed-form block LOO-CV (Rasmussen & Williams, 2006, eq. 5.12) evaluates
predictive performance at each NOAA site without refitting.

```{r loo}
loo <- loo_cv(model)
sum_loo <- loo_summary(loo, r = 100)

# Total log predictive density
sum_loo$total_lpd

# 100-year return level RMSE
sum_loo$rl_rmse
```

## Custom Data Sources

The `source_params` argument generalizes the framework beyond NOAA/ADCIRC.
Each entry maps a data source name to its parameter indices in the joint
model:

```{r custom, eval = FALSE}
# Example: 3 sources, each observing 3 GEV parameters
dat <- load_data(df, source_params = list(
  gauge    = 1:3,
  satellite = 4:6,
  reanalysis = 7:9
))

# Fit a 9-dimensional joint model
model <- fit_spatial_model(stage1, dat, W_tap, D)
```

## References

White, B. N., Blanton, B., Luettich, R., & Smith, R. L. Fusing Sparse
Observations and Dense Simulations for Spatial Extreme Value Analysis:
Application to U.S. Coastal Sea Levels. *In preparation.*

Russell, B. T., Cooley, D. S., Porter, W. C., Reich, B. J., & Heald, C. L.
(2019). Data fusion for environmental and human health applications.
*International Statistical Review*, 87, S261--S287.

Rasmussen, C. E. & Williams, C. K. I. (2006). *Gaussian Processes for Machine
Learning*. MIT Press.
